The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-24 paper/ms.tex
Update ack
Small changes throughout

\subsection{Cartoon Effect of Midplane Error} \label{ssec:cartoon}
We present a cartoon approximation in Figure~\ref{fig:cartoon} to show how an
inaccurate determination of the midplane leads to a phase-dependence of the
actions calculated from an orbit specified by a phase-space point and an
assumed potential model. The $y$-axis corresponds to the vertical height of
the orbit and the $x$-axis orbital phase. The solid gray line indicates the
``true'' orbit of the star; that is, its orbit as it oscillates around the
true midplane. The dashed gray line is an erroneous midplane assigned to the
potential model which an observer uses to integrate the orbit of the star and
hence estimate its actions. The model potential is otherwise identical to the
one in which the star is actually moving.

Now suppose this observer makes a measurement of the true orbit at the purple
point or the green point (i.e. at two different orbital phases). Then the
purple and green lines correspond to the orbits that the observer would
compute for each point based on the potential model with the erroneous
midplane. In action space, this would correspond to a different value of $J_z$
for the blue and pink points. In this way, assuming the wrong coordinate
system induces a phase-dependence in the actions estimated for the star, which
in the correct potential (in this example, the one with the correct midplane)
should be phase-independent.

@@

There is a slight complication in the formula for computing $\Delta J_R/J_R$
and $\Delta J_{\phi}/J_{\phi}$. In a moment we will consider the effect of an inaccuracy
in the distance from the Sun to the Galactic center on actions. However, this
inaccuracy induces an error in $x$, or $\Delta x$. This is not the same as an
inaccuracy in $\Delta R$, but they are related by the following argument. An
error $\Delta x$ will induce an erroneous radius $R_{\text{err}}$ related by
the formula,
\beq
(x+\Delta x)^2 + y^2 = R_{\text{err}}^2\text{.}
\eeq
Keeping only terms to first order in $\Delta x$, we have that,
\beq
\begin{split}
R_{\text{err}}^2 &= R^2 - 2 R \cos{\theta} \Delta x \\
\implies \Delta R &\equiv \abs{R_{\text{err}} - R} = \abs{\cos{\theta}} \Delta x\text{.}
\end{split}
\eeq
Averaging over the circle, we therefore have that
\beq
\avg{\Delta R} = \frac{2}{\pi} \Delta x\text{.}
\eeq
Therefore, the error in $J_R$ and $J_{\phi}$ are related to $\Delta x$ by,
\beq\label{eq:Jr_epi_dx}
\begin{split}
\frac{\Delta J_R}{J_R} &= \frac{4 \Delta x}{\pi A_R} \\
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{2 \Delta x}{\pi R}\text{.}
\end{split}
\eeq

@@

We begin by considering the thick-disk orbit. Suppose that we could observe
the thick-disk star's phase-space position at many different times (and hence
different orbital phases), but used a coordinate system in which the midplane
is systematically offset in height by $100\,\pc$ from its actual location ---
i.e. we subtract the vector $(0, 0, 100)\,\pc$ from each position in the
orbit. This corresponds to an observer physically located at e.g. the position
$(8, 0, 0)\,\kpc$ in the coordinate system of the true potential, but
erroneously thinking they are located at $(8, 0, 0.1)\,\kpc$.

@@

Starting from each observation of the star's phase-space position that such an
(extremely long-lived) observer makes every $\Myr$, we then compute the
actions as outlined above using an integration from that starting point in
phase space carried out using the true potential but specifying the star's
present-day position using the systematically offset coordinate system.
Essentially we are shifting the integrated orbit and then reintegrating at
each point along the shifted orbit, where the shifted orbit is the observed
orbit. The erroneous computed actions for each phase-space starting point are
shown for the first $\Gyr$ of the orbit in the {\em upper} panels
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $5\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To our extremely long-lived observer
this would appear as time-dependence in the actions, which appear to oscillate
around their value using the correct coordinate system even though in this
example the observer is using the correctly constructed best-fit static,
axisymmetric potential. The relative size of the phase variation in each
action depends on the direction of the systematic offset as well as the true
values of the actions (i.e. the type of orbit). In reality we will have one
measurement of the phase-space position to work with, in which case the
determination of the orbital phase in $z$ or $R$ is degenerate with the degree
of systematic offset in that coordinate (see Figure~\ref{fig:cartoon}).
Additionally, in searching for migrated members of known open clusters, we
would compute actions for stars with a wide range of orbital phases and
locations in the Galaxy. In the following we therefore quote percentile ranges
for the possible values computed for each action as a proxy for the effect of
these systematic errors in the coordinate system.

@@

For a systematic offset in $x$ (or Galactic center distance), the
95\textsuperscript{th} minus 5\textsuperscript{th} percentiles are $6.9$,
$47$, and $0.71\,\actunit$ for $J_R$, $J_{\phi}$ and $J_z$, respectively. These are
fractional errors of $21\%$, $3.1\%$ and $3.1\%$, respectively.

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the observed values of
$J_z$ for the thick-disk orbit ({\em upper panel}) and thin-disk orbit ({\em
lower panel}) assuming a $z$ offset of $100\,\pc$ (i.e. the {\em right} panel
of Figure~\ref{fig:one_orbit_wrong_ref}). The true value of $J_z$ is plotted
as a vertical dashed line. Here we see that the errors in $J_z$ induced by a
systematic offset in $z$ are non-Gaussian and bimodal. Furthermore, neither of
the modes lie at zero. In other words, neither of the modes of the
distribution of observed $J_z$ values are centered on the true $J_z$ (i.e. the
$J_z$ computed with no offset in $z$). In the case of the thin-disk orbit
({\em lower panel}), we see that, in addition to the prior complications, the
error distribution is not even centered on null. This comes about when the
midplane error is $\gtrsim z_{\text{max}}$, and can be understood by realizing
that in such a case the midplane offset always acts to increase the value of
$J_z$, resulting in a biased estimate of $J_z$.

The intuition behind the shape of Figure~\ref{fig:Jz_hist} is relatively
simple. Consider first the thick-disk orbit ({\em top} panel), where the
offset in $z$ is much less than $z_{\text{max}}$ of the orbit. The peaks in
the distribution correspond to the turning points of the orbit (or points of
maximum vertical amplitude), where $v_z \sim 0$ and where the star spends
most of it orbit. This is why the distribution is peaked on these values. For
the thin-disk orbit ({\em bottom} panel), the offset in $z$ is comparable to
$z_{\text{max}}$. Now, there will be some points in the orbit where $v_z = 0$
and $z=0$ (in the observed, erroneous coordinate system). At these points, the
computed $J_z$ will vanish. The asymmetry and systematic offset then comes
about because of the constraint that $J_z \geq 0$.\footnote{This is similar to
arguments in cosmology for why gravity produces non-Gaussianity in the density
field, since the density cannot become negative but it can grow arbitrarily
large.}

@@

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report one half the
95\uth{} minus 5\uth{} percentile divided by the true action value ($\Delta
J_i/J_i$) for the three different fiducial orbits in Table \ref{tab:orbits}.
The thick-disk (\thickcolor) orbit is the one from
Figure~\ref{fig:one_orbit_wrong_ref}, but we also now consider the effect on
the action determined for the thin-disk (\thincolor) and halo (\halocolor)
orbits.

@@

The {\em upper right} panel of Figure~\ref{fig:many_orbit_wrong_ref} shows
that the fractional error in $J_z$ is more exaggerated for more planar
(disk-like) orbits. For the thin-disk orbit, a systematic offset of
$\sim30\,\pc$ in the $z$ coordinate already results in $100\%$ deviations in
the actions. The required offset for $100\%$ deviation is $\sim225\,\pc$ for
the thick-disk orbit. The halo orbit is relatively resistant to errors in the
midplane, with only $\sim25\%$ error in $J_z$ out to a midplane offset of
$500\,\pc$.

@@

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (red, cyan, green, respectively) in Figure~\ref{fig:dJz_fun_Jz}. For
a thin disk-like orbit ($J_z\sim0.5\,\actunit$), even a $10\,\pc$ offset in
$z$ is enough to induce a $\sim50\%$ error in $J_z$. For larger values of
$J_z$, the fractional errors are suppressed, but the induced error can still
be large depending on how great the $z$ offset is.

@@

\subsection{Galactic Orientation}
Second, one must define the orientation of the Galaxy about the center. This
was defined in 1958 by the IAU subcomission 33b \citep{1960MNRAS.121..123B} by
defining the location of the Galactic center and the north Galactic pole.
However, there is growing evidence that the true midplane is tilted when
considering stars \citep{2014ApJ...797...53G, 2016ARAA..54..529B}, though not
with H~\textsc{ii} \citep{2019ApJ...871..145A}.

\subsection{Solar Height}
Thirdly, one must define The Sun's vertical distance from the Galactic
midplane, which can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars) The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from {\em Gaia} DR2
placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another strategy is
to use the cold gas or H~\textsc{ii} regions in the disk to define the Galactic
midplane, leading to slightly different values ($\sim 5\,\pc$) for the Sun's
relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review
of these measurements is given by \citet{2016ARAA..54..529B}.

@@

We now focus our attention on understanding the intrinsic variations of the
midplane across the disk in two sets of complementary simulations. We will
briefly discuss a rough estimate of the midplane variation derived from the
observed velocity distribution of the Milky Way, but defer a detailed
treatment to future work.

@@

\subsection{Description of Sagittarius Simulaton} \label{ssec:sag_sim}
In addition to the cosmological zoom-in, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modelled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modelled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modelled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$.

For the disk and bulge components, a softening length of $30\,\pc$ is used
whereas for the halo a softening length of $60\,\pc$ is used. For Sagittarius,
the softening length for the dark matter and the stars is $60$ and $40\,\pc$,
respectively.

\subsection{The Local Midplane} \label{ssec:local_midplane}
Using the three simulations, we determine the local midplane that an observer
might measure if they were situated in each of these galaxies as a function of
azimuth at the solar circle. Starting from the coordinate system described in
the previous section, which is approximately aligned so that the
$z$-coordinate is perpendicular to the disk plane, we place our imaginary
observer at $z=0$ and a solar radius of $8.2\,\kpc$ and vary the azimuth
between $0<\phi<2\pi$\footnote{Our very long-lived observer also has warp
drive.}. At each value of $\phi$ we then compute the median $z$ for stars
within a cylinder of radius $0.5\,\kpc$ and height $1\,\kpc$ perpendicular to
the fiducial disk. We then re-define this median $z$ as the new midplane of
the cylinder, re-select stars, and iterate until the median $z$ value
converges. We find that only $10$ iterations of this procedure are necessary
for convergence. The resulting median $z$ is taken to be what our observer
would measure as the local Galactic midplane at each $\phi$. This procedure
assumes perfect density estimation, and therefore perfect corrections for
extinction, within the cylinder defining the ``solar neighborhood.'' Imperfect
extinction correction is likely to increase the amplitude of the estimated
fluctuations in $z$. To account for the effect of particle noise, we
bootstrap-resample stars within a cylinder of stars with height $2\,\kpc$ and
the same radius $1000$ times and determine the $1\,\sigma$ error bars. This
bootstrap resampling is performed by first selecting all stars within a height
of $2\,\kpc$, resampling that selection, and then repeating the $10$
iterations.

@@

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for the toy simulation of a Sagittarius encounter
\citep{2018MNRAS.481..286L}. Again we have subtracted a best fit curve of the
form $\text{A} \sin{\left(\phi + \text{B}\right)} + \text{C}$, with the values
for $\text{A}$ being $9.5, 2.5, -212, -394\,\pc$, for $\text{B}$ being
$0.0013, 0.0007, -0.63, -1.00\,\text{rad}$ and $\text{C}$ being $2.6, -7.8,
-65, -53\,\pc$. Each of these values are for each successive timestep. In the
first two panels, we see that the midplane is relatively flat in the inner
galaxy. But additional encounters drive strong midplane variation. In the
third panel, we see a strong $m=2$ mode develop, consistent with the
$R=8\,\kpc$ panel of Figure~17 in \citet{2018MNRAS.481..286L}.

\begin{figure*}
\begin{center}
\includegraphics[width=7in]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma
$ error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

@@

An interesting, qualitative calculation can be made based on the velocity
variations seen in e.g. \citet{2019arXiv190209569F}. The velocity variations
in this work (admittedly found with respect to $J_{\phi}$ or radius) are
$\sim5\text{--}10\,\kms$. The vertical frequency of the thin-disk and
thick-disk orbits are $\sim 0.09\,1/\Myr$ and $\sim0.06\,1/\Myr$,
respectively. By dimensional analysis, we therefore expect the midplane
offsets to be of order $\sim 57\text{--}170\,\pc$. We stress that this is a
rough calculation, but it shows that already we see evidence in the data from
velocities for midplane offsets on the order of what we see in both sets of
simulations.

@@

Suppose an open cluster of mass $m_c$ on an orbit with actions $\bm{J}_c$ is
disrupted. We would like to determine which stars with complete 6D phase-space
positions, for example in the {\em Gaia} dataset,\footnote{This is likely to
be more successful with DR3/4 than DR2, since upcoming data releases will
contain a far greater number of stars with radial velocities --- though
targeted, follow-up radial velocity surveys may allow this program to be
performed sooner.} are likely to be ejected members from that cluster. One
could generate an initial catalogue of candidate members by integrating the
cluster's orbit and selecting stars close to that orbit, accounting for the
fact that the stream of ejected stars will not exactly follow the orbit of the
cluster \citep[e.g.][]{2011MNRAS.413.1852E}.

One could then make a further selection by requiring that the action of each
field star $i$ be close to the cluster's actions to within some bound, i.e.
$\bm{J}_i \in V_e$, where $V_e$ is a volume centered on $\bm{J}_c$ and has size
comparable to the intrinsic action-space dispersion of the disrupting
cluster. This dispersion can be estimated as \citep[Section~8.3.3
of][]{2008gady.book.....B}
\beq \label{eq:action_disp}
\frac{\Delta J_i}{J_i} \sim \left(\frac{m_c}{M}\right)^{1/3}\text{,}
\eeq
where $m_c$ is the mass of the cluster and $M$ is the mass enclosed by the
cluster's orbit around the Galaxy, in this case assumed to be the mass inside
the Solar circle ($\sim 10^{11}\Msun$). Stars with actions within $V_e$ would
then be considered cluster members.

@@

\begin{itemize}
\item We demonstrated that an inaccuracy in the Galactocentric coordinate
system induces time evolution in the {\em observed} actions
(Figures~\ref{fig:cartoon}\&\ref{fig:one_orbit_wrong_ref}).

>>>>

\subsection{Cartoon Effect of Midplane Error} \label{ssec:cartoon}
We present a cartoon approximation in Figure~\ref{fig:cartoon} to show how an
inaccurate determination of the midplane leads to a phase-dependence of the
actions calculated from an orbit specified by a phase-space point and an
assumed potential model. The $y$-axis corresponds to the vertical height of
the orbit and the $x$-axis orbital phase. The solid gray curve indicates the
``true'' orbit of the star; that is, its orbit as it oscillates around the
true midplane. The dashed gray line is an erroneous midplane assigned to the
potential model which an observer uses to integrate the orbit of the star and
hence estimate its actions. The model potential is otherwise identical to the
one in which the star is actually moving.

Now suppose this observer makes a measurement of the true orbit at the purple
point or the green point (i.e. at two different orbital phases). Then the
purple and green curves correspond to the orbits that the observer would
compute for each point based on the potential model with the erroneous
midplane. In action space, this would correspond to a different value of $J_z$
for the purple and green points. In this way, assuming the wrong coordinate
system induces a phase-dependence in the actions estimated for the star, which
in the correct potential (in this example, the one with the correct midplane)
should be phase-independent.

@@

There is a slight complication in the formula for computing $\Delta J_R/J_R$
and $\Delta J_{\phi}/J_{\phi}$. In a moment we will consider the effect of an inaccuracy
in the distance from the Sun to the Galactic center on actions. However, this
inaccuracy induces an error in $x$, or $\Delta x$. This is not the same as an
inaccuracy in $\Delta R$, but they are related by the following argument. An
error $\Delta x$ will induce an erroneous radius $R_{\text{err}}$ related by
the formula,
\beq
(x+\Delta x)^2 + y^2 = R_{\text{err}}^2\text{.}
\eeq
Keeping only terms to first order in $\Delta x$, we have that,
\beq
\begin{split}
R_{\text{err}}^2 &= R^2 - 2 R \cos{\theta} \Delta x \\
\implies \Delta R &\equiv \abs{R_{\text{err}} - R} = \abs{\cos{\theta}} \Delta x\text{.}
\end{split}
\eeq
Averaging over the circle, we therefore have that,
\beq
\avg{\Delta R} = \frac{2}{\pi} \Delta x\text{.}
\eeq
Therefore, the error in $J_R$ and $J_{\phi}$ are related to $\Delta x$ by,
\beq\label{eq:Jr_epi_dx}
\begin{split}
\frac{\Delta J_R}{J_R} &= \frac{4 \Delta x}{\pi A_R} \\
\frac{\Delta J_{\phi}}{J_{\phi}} &= \frac{2 \Delta x}{\pi R}\text{.}
\end{split}
\eeq

@@

We begin by considering the thick-disk orbit. Suppose an immortal observer who can observe
the thick-disk star's phase-space position at many different times (and hence
different orbital phases), but used a coordinate system in which the midplane
is systematically offset in height by $100\,\pc$ from its actual location ---
i.e. we subtract the vector $(0, 0, 100)\,\pc$ from each position in the
orbit. This corresponds to the immortal observer physically located at e.g. the position
$(8, 0, 0)\,\kpc$ in the coordinate system of the true potential, but
erroneously thinking they are located at $(8, 0, 0.1)\,\kpc$.

@@

Starting from each observation of the star's phase-space position that such an
immortal observer makes every $\Myr$, we then compute the
actions as outlined above using an integration from that starting point in
phase space carried out using the true potential but specifying the star's
present-day position using the systematically offset coordinate system.
Essentially we are shifting the integrated orbit and then reintegrating at
each point along the shifted orbit, where the shifted orbit is the observed
orbit. The erroneous computed actions for each phase-space starting point are
shown for the first $\Gyr$ of the orbit in the {\em upper} panels
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $5\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the immortal observer
this would appear as time-dependence in the actions, which appear to oscillate
around their value using the correct coordinate system even though in this
example the observer is using the correctly constructed best-fit static,
axisymmetric potential. The relative size of the phase variation in each
action depends on the direction of the systematic offset as well as the true
values of the actions (i.e. the type of orbit). In reality we will have one
measurement of the phase-space position to work with, in which case the
determination of the orbital phase in $z$ or $R$ is degenerate with the degree
of systematic offset in that coordinate (see Figure~\ref{fig:cartoon}).
Additionally, in searching for migrated members of known open clusters, we
would compute actions for stars with a wide range of orbital phases and
locations in the Galaxy. In the following we therefore quote percentile ranges
for the possible values computed for each action as a proxy for the effect of
these systematic errors in the coordinate system.

@@

For a systematic offset in $x$ (or Galactic center distance), the
95\textsuperscript{th} minus 5\textsuperscript{th} percentiles are $6.9$,
$47$, and $0.71\,\actunit$ for $J_R$, $J_{\phi}$ and $J_z$, respectively.
These are fractional errors of $21\%$, $3.1\%$ and $3.1\%$, respectively ---
despite only a $1.2\%$ error in the distance to the Galactic center.

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the observed values of
$J_z$ for the thick-disk orbit ({\em upper panel}) and thin-disk orbit ({\em
lower panel}) assuming a $z$ offset of $100\,\pc$ (i.e. the {\em upper right} panel
of Figure~\ref{fig:one_orbit_wrong_ref}). The true value of $J_z$ is plotted
as a vertical dashed line. Here we see that the errors in $J_z$ induced by a
systematic offset in $z$ are non-Gaussian and bimodal. Furthermore, neither of
the modes lie at zero. In other words, neither of the modes of the
distribution of observed $J_z$ values are centered on the true $J_z$ (i.e. the
$J_z$ computed with no offset in $z$). In the case of the thin-disk orbit
({\em lower panel}), we see that, in addition to the prior complications, the
error distribution is not even centered on null. This comes about when the
midplane error is $\gtrsim z_{\text{max}}$.

We now describe the intuition behind the shape of Figure~\ref{fig:Jz_hist}.
Consider first the thick-disk orbit ({\em top} panel), where the
offset in $z$ is much less than $z_{\text{max}}$ of the orbit. The peaks in
the distribution correspond to the turning points of the orbit (or points of
maximum vertical amplitude), where $v_z \sim 0$ and where the star spends
most of it orbit. This is why the distribution is peaked on these values. For
the thin-disk orbit ({\em bottom} panel), the offset in $z$ is comparable to
$z_{\text{max}}$. Now, there will be some points in the orbit where $v_z = 0$
and $z=0$ (in the observed, erroneous coordinate system). At these points, the
computed $J_z$ will vanish. The asymmetry and systematic offset then comes
about because of the constraint that $J_z \geq 0$.\footnote{This is similar to
arguments in cosmology for why gravity produces non-Gaussianity in the density
field, since the density cannot become negative but it can grow arbitrarily
large.}

@@

We now repeat the same procedure as in Figure~\ref{fig:one_orbit_wrong_ref}
but for systematic offsets between $0$ and $500\,\pc$ in the $z$ and $x$
components. In Figure~\ref{fig:many_orbit_wrong_ref}, we report one half the
95\uth{} minus 5\uth{} percentile divided by the true action value ($\Delta
J_i/J_i$) for the three different fiducial orbits in Table~\ref{tab:orbits}.
The thick-disk (\thickcolor) orbit is the one from
Figure~\ref{fig:one_orbit_wrong_ref}, but we also now consider the effect on
the action determined for the thin-disk (\thincolor) and halo (\halocolor)
orbits.

@@

The {\em upper right} panel of Figure~\ref{fig:many_orbit_wrong_ref} shows
that the fractional error in $J_z$ is more exaggerated for more planar
(disk-like) orbits. For the thin-disk orbit, a systematic offset of
$\sim15\,\pc$ in the $z$ coordinate already results in $25\%$ deviations in
the actions. The required offset for $25\%$ deviation is $\sim120\,\pc$ for
the thick-disk orbit. The halo orbit is relatively resistant to errors in the
midplane, with only $\sim15\%$ error in $J_z$ out to a midplane offset of
$500\,\pc$.

@@

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, cyan, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For a thin disk-like orbit
($J_z\sim0.5\,\actunit$), even a $10\,\pc$ offset in $z$ is enough to induce a
$\sim25\%$ error in $J_z$. For larger values of $J_z$, the fractional errors
are suppressed, but the induced error can still be large depending on how
great the $z$ offset is. We also plot the epicylic prediction for $\Delta J_z
/ J_z$ from Equation~\ref{eq:actions_epi_error} as dashed lines for each $z$
offset. We find that the epicyclic approximation matches the numerical
estimate quite well.

@@

\subsection{Galactic Orientation}
Second, one must define the orientation of the Galaxy about the center. This
was defined in 1958 by the IAU subcomission 33b \citep{1960MNRAS.121..123B} by
defining the coordinates of the Galactic center in B1950 coordinates as
(17:42:26.6, -28:55:00) and the north Galactic pole as (12:49:00, +27:24:00).
However, there is growing evidence that the true midplane is tilted when
considering stars \citep{2014ApJ...797...53G, 2016ARAA..54..529B}, though not
with H~\textsc{ii} \citep{2019ApJ...871..145A}.

\subsection{Solar Height}
Thirdly, one must define the Sun's vertical distance from the Galactic
midplane, which can be determined by identifying where the stellar density
and velocities reach a maximum (effectively the median height of all disk
stars). The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from {\em Gaia} DR2
placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another strategy is
to use the cold gas or H~\textsc{ii} regions in the disk to define the Galactic
midplane, leading to slightly different values ($\sim 5\,\pc$) for the Sun's
relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em Gaia} review
of these measurements is given by \citet{2016ARAA..54..529B}.

@@

We now consider a second source of systematic offsets in the Galactocentric
coordinate system --- intrinsic variations of the midplane across the disk. We
study this in two sets of simulations of Milky Way-mass galaxies. We will also
briefly discuss a rough estimate of the midplane variation derived from the
observed velocity distribution of the Milky Way, but defer a detailed
treatment to future work.

@@

\subsection{Description of Sagittarius Simulaton} \label{ssec:sag_sim}
In addition to the cosmological zoom-in, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modelled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modelled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modelled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$.

The mass resolution of the simulation is $2.6\times10^4$, $1.2\times10^4$, and
$1.0\times10^4$ for the dark matter, disk, and bulge components, respectively.
For the disk and bulge components, a softening length of $30\,\pc$ is used
whereas for the halo a softening length of $60\,\pc$ is used. For Sagittarius,
the softening length for the dark matter and the stars is $60$ and $40\,\pc$,
respectively.

\subsection{The Local Midplane} \label{ssec:local_midplane}
Using the two sets of simulations, we determine the local midplane that an observer
might measure if they were situated in each of these galaxies as a function of
azimuth at the solar circle. Starting from the coordinate system described in
the previous section, which is approximately aligned so that the
$z$-coordinate is perpendicular to the disk plane, we place our imaginary
observer at $z=0$ and a solar radius of $8.2\,\kpc$ and vary the azimuth
between $0<\phi<2\pi$.\footnote{The immortal observer also has warp
drive.} At each value of $\phi$ we then compute the median $z$ for stars
within a cylinder of radius $0.5\,\kpc$ and height $1\,\kpc$ perpendicular to
the fiducial disk. We then re-define the new midplane of the cylinder to be the median $z$, re-select stars, and iterate until the median $z$ value
converges. We find that only $10$ iterations of this procedure are necessary
for convergence. The resulting median $z$ is taken to be what our observer
would measure as the local Galactic midplane at each $\phi$. This procedure
assumes perfect density estimation, and therefore perfect corrections for
extinction, within the cylinder defining the ``solar neighborhood.'' Imperfect
extinction correction is likely to increase the amplitude of the estimated
fluctuations in $z$. To account for the effect of particle noise, we
bootstrap-resample stars within a cylinder of stars with height $2\,\kpc$ and
the same radius $1000$ times and determine the $1\,\sigma$ error bars by repeating the midplane determination with those reselection of stars.

@@

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for the toy simulation of a Sagittarius encounter
\citep{2018MNRAS.481..286L}. Again we have subtracted a best fit curve of the
form $\text{A} \sin{\left(\phi + \text{B}\right)} + \text{C}$, with the values
for $\text{A}$ being $9.5$, $2.5$, $-212$, and $-394\,\pc$, for $\text{B}$ being
$0.0013$, $0.0007$, $-0.63$, and $-1.00\,\text{rad}$ and $\text{C}$ being $2.6$, $-7.8$,
$-65$, and $-53\,\pc$. Each of these values are for each successive timestep. In the
first two panels, we see that the midplane is relatively flat in the inner
galaxy. But additional encounters drive strong midplane variation. In the
third panel, we see a strong $m=2$ mode develop, consistent with the
$R=8\,\kpc$ panel of Figure~17 in \citet{2018MNRAS.481..286L}.

\begin{figure*}
\begin{center}
\includegraphics[width=7in]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). All calculations are done at cosmological redshift $\z =0$. The local midplane is determined at a
position $\phi$ by taking the median height of all stars within $R=0.5\,\kpc$
and $z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma
$ error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

@@

An interesting, qualitative calculation can be made based on the velocity
variations seen in e.g. \citet{2019arXiv190209569F}. The velocity variations
in this work (admittedly found with respect to $J_{\phi}$ or radius) are
$\sim5\text{--}10\,\kms$. The vertical frequency of the thin-disk and
thick-disk orbits are $\sim 0.09\,\Myr^{-1}$ and $\sim0.06\,\Myr^{-1}$,
respectively. By dimensional analysis, we therefore expect the midplane
offsets to be of order $\sim 57\text{--}170\,\pc$. We stress that this is a
rough calculation, but it shows that already we see evidence in the data from
velocities for midplane offsets on the order of what we see in both sets of
simulations.

@@

Suppose an open cluster of mass $m_c$ on an orbit with actions $\bm{J}_c$ is
disrupted. We would like to determine which stars with complete 6D phase-space
positions, for example in the {\em Gaia} dataset, are likely to be ejected members from that cluster.\footnote{This is likely to
be more successful with DR3/4 than DR2, since upcoming data releases will
contain a far greater number of stars with radial velocities --- though
targeted, follow-up radial velocity surveys may allow this program to be
performed sooner.} One
could generate an initial catalogue of candidate members by integrating the
cluster's orbit and selecting stars close to that orbit, accounting for the
fact that the stream of ejected stars will not exactly follow the orbit of the
cluster \citep[e.g.][]{2011MNRAS.413.1852E}.

One could then make a further selection by requiring that the action of each
field star $i$ be close to the cluster's actions to within some bound, i.e.
$\bm{J}_i \in V_e$, where $V_e$ is a volume centered on $\bm{J}_c$ and has size
comparable to the intrinsic action-space dispersion of the disrupting
cluster. This dispersion can be estimated as \citep[Section~8.3.3][]{2008gady.book.....B},
\beq \label{eq:action_disp}
\frac{\Delta J_i}{J_i} \sim \left(\frac{m_c}{M}\right)^{1/3}\text{,}
\eeq
where $m_c$ is the mass of the cluster and $M$ is the mass enclosed by the
cluster's orbit around the Galaxy, in this case assumed to be the mass inside
the Solar circle ($\sim 10^{11}\Msun$). Stars with actions within $V_e$ would
then be considered cluster members.

@@

\begin{itemize}
\item We demonstrated that an inaccuracy in the Galactocentric coordinate
system induces time evolution in the {\em observed} actions
(Figures~\ref{fig:cartoon}~and~\ref{fig:one_orbit_wrong_ref}). We interpret
this time evolution as an error in the computed action
(Figure~\ref{fig:many_orbit_wrong_ref}).
